Denotational Semantics of Gradual Typing using Synthetic Guarded Domain Theory

### 2024-01-27 paper-new/intro.tex
Revise sections 1 and 2

Our goal in this work is to provide a reusable semantic framework for
gradually typed languages that can be used to prove the aforementioned
properties: graduality and type-based reasoning. \max{TODO: say more here}

@@

The standard alternative to domain theory that scales to essentially
arbitrary self-referential definitions is \emph{step-indexing} or its
synthetic form of \emph{guarded recursion}.  \max{todo: adapt the next
  paragraph to fit in with this one} A series of works
\cite{new-ahmed2018, new-licata-ahmed2019, new-jamner-ahmed19}
developed step-indexed logical relations models of gradually typed
languages based on operational semantics. Unlike classical domain
theory, such step-indexed techniques are capable of modeling
higher-order store and runtime-extensible dynamic types
\cite{appelmcallester01,ahmed06,neis09,new-jamner-ahmed19}. However,
their proof developments are highly repetitive and technical, with
each development formulating a logical relation from first-principles
and proving many of the same tedious lemmas without reusable
mathematical abstractions. Our goal in the current work is to extract
these reusable mathematical principles from these explicit
step-indexed to make formalization of realistic gradual languages
tractible.

@@

\max{the stuff below is out of date}
An alternative approach, which we investigate in this paper, is provided by
\emph{synthetic guarded domain theory}\cite{birkedal-mogelberg-schwinghammer-stovring2011}.
The techniques of synthetic guarded domain theory allow us to internalize the
step-indexed reasoning normally required in logical relations proofs of graduality,
ultimately allowing us to specify the logical relation in a manner that looks nearly
identical to a typical, non-step-indexed logical relation.

In this paper, we report on work we have done towards mechanizing proofs of graduality
and soundness of the equational theory of cast calculi using the techniques of SGDT.
We take a step toward mechanization in the Agda proof assistant by describing a compositional
denotational semantics for gradual typing in a setting where the steps taken by terms are tracked.

Our longer-term goal is to mechanize these proofs in a reusable way, thereby providing a framework
to use to more easily and conveniently prove that existing languages satisfy graduality and have
sound equational theories. Moreover, the aim is for designers of new languages to utilize the
framework to facilitate the design of new provably-correct gradually-typed languages with more
complex features.

@@

The main contribution of this work is a categorical and denotational
semantics for gradually typed langauges that models not just the term
language but the graduality property as well.
\begin{enumerate}
\item First, we give a simple abstract categorical model of GTT using CBPV double categories.
\item Next, we modify this semantics to develop reflexive graph- and
  double categorical models that abstract over the details of
  step-indexed models, and provide a method for constructing such models.
\item We instantiate the abstract construction to provide a concrete
  semantics in informal guarded type theory which supports the
  addition of arbitrary algebraic effects.
\item We prove that the resulting denotational model is
  \emph{adequate} for the graduality property: a closed term precision
  $M \ltdyn N : \nat$ has the expected semantics, that $M$ errors or
  $M$ and $N$ have the same extensional behavior.
\end{enumerate}

Our main contribution is a compositional denotational semantics for step-aware gradual typing,

analogous to Gradual Type Theory but now in the ``intensional" setting.
In parallel, we are developing a reusable framework in
Guarded Cubical Agda for developing machine-checked proofs of graduality of a cast calculus.
To demonstrate the feasibility and utility of our approach, we are applying the framework
to prove graduality for the simply-typed gradual lambda calculus.
% Along the way, we have developed an ``intensional" theory of graduality that is of
% independent interest.

\subsection{Outline of Remainder of Paper}

\max{this is entirely out of date, update later}

% In Section \ref{sec:overview}, we give an overview of the gradually-typed lambda
% calculus and the graduality theorem.
%
In Section \ref{sec:technical-background}, we provide technical background on synthetic guarded domain theory.
% 
In Section \ref{sec:gtlc-terms}, we formally introduce the gradually-typed cast calculus
for which we will prove graduality. We give a semantics to the terms using
synthetic guarded domain theory.
% Important here are the notions of syntactic
% type precision and term precision. For reasons we describe below, we
% introduce two related calculi, one in which the stepping behavior of terms is
% implicit (an \emph{extensional} calculus, $\extlc$), and another where this behavior
% is made explicit (an \emph{intensional} calculus, $\intlc$).
%

In Section \ref{sec:gtlc-precision}, we define the term precision ordering for
the gradually-typed lambda calculus and describe our approach to assigning a
denotational semantics to this ordering.
This approach builds on the semantics constructed in the previous section,
but extending it to the term ordering presents new challenges.

% In Section \ref{sec:domain-theory}, we define several fundamental constructions
% internal to SGDT that will be needed when we give a denotational semantics to
% the gradual lambda calculus.
%This includes the Lift monad, predomains and error domains.
%
% In Section \ref{sec:semantics}, we define the denotational semantics for the
% intensional gradually-typed lambda calculus using the domain theoretic
% constructions defined in the previous section.
%
In Section \ref{sec:graduality}, we outline in more detail the proof of graduality for the
gradual lambda calculus.
%
In Section \ref{sec:discussion}, we discuss the benefits and drawbacks to our approach in comparison
to the traditional step-indexing approach, as well as possibilities for future work.

\subsection{Overview of Results}\label{sec:overview}

% This section used to be part of the intro.
% \subsection{Proving Graduality in SGDT}
% TODO This section should probably be moved to after the relevant background has been introduced?

% TODO introduce the idea of cast calculus and explicit casts?

In this paper, we will utilize SGDT techniques to prove graduality for a particularly
simple gradually-typed cast calculus, the gradually-typed lambda calculus.
Before stating the graduality theorem, we introduce some definitions related to gradual typing.

% Cast calculi
% TODO move this to earlier?
Gradually typed languages are generally elaborated to a \emph{cast calculus}, in which run-time type checking
that is made explicit. Elaboration inserts \emph{type casts} at the boundaries between static and dynamic code.
In particular, cast insertion happens at the elimination forms of the language (e.g., pattern matching, field reference, etc.).
The original gradually typed language that is elaborated to a cast calculus is called the \emph{surface language}.

% Up and down casts
In a cast calculus, there is a relation $\ltdyn$ on types such that $A \ltdyn B$ means that
$A$ is a \emph{more precise} type than $B$.
There a distinguished type $\dyn$, the \emph{dynamic type}, with the property that $A \ltdyn\, \dyn$ for all $A$.
The parts of the code that are dynamically typed will be given type $\dyn$ in the cast calculus.
%
If $A \ltdyn B$, a term $M$ of type $A$ may be \emph{up}casted to $B$, written $\up A B M$,
and a term $N$ of type $B$ may be \emph{down}casted to $A$, written $\dn A B N$.
Upcasts always succeed, while downcasts may fail at runtime, resulting in a type error.
Cast calculi have a term $\mho$ representing a run-time type error for any type $A$.

% Syntactic term precision
We also have a notion of \emph{syntactic term precision}.
If $A \ltdyn B$, and $M$ and $N$ are terms of type $A$ and $B$ respectively, we write
$M \ltdyn N : A \ltdyn B$ to mean that
$M$ is ``syntactically more precise'' than $N$, or equivalently, $N$ is 
``more dynamic'' than $M$. Intuitively, this models the situation where $M$ and $N$
behave the same except that $M$ may error more.
Term precision is defined by a set of axioms that capture this intuitive notion.
The specific rules for term precision depend on the specifics of the language, but
many of the rules can be derived from the typing rules in a straightforward way.
% Cast calculi also have a term $\mho$ representing a run-time type error for any type $A$,
% and since
Since $\mho$ always errors, there is a term precision rule stating that $\mho \ltdyn M$ for all $M$.

% GTLC
The gradually-typed lambda calculus is the usual simply-typed lambda calculus with a dynamic
type $\dyn$ such that $A \ltdyn\, \dyn$ for all types $A$, as well as upcasts and downcasts
between any types $A$ and $B$ such that $A \ltdyn B$. The complete definition will be provided in
Section \ref{sec:gtlc-precision}.

With these definitions, we can state the graduality theorem for the gradually-typed lambda calculus.

% \begin{theorem}[Graduality]
%   If $M \ltdyn N : \nat$, then either:
%   \begin{enumerate}
%     \item $M = \mho$
%     \item $M = N = \zro$
%     \item $M = N = \suc n$ for some $n$
%     \item $M$ and $N$ diverge
%   \end{enumerate}
% \end{theorem}

\begin{theorem}[Graduality]
  Let $\cdot \vdash M \ltdyn N : \nat$. 
  If $M \Downarrow v_?$ and $N \Downarrow v'_?$, then either $v_? = \mho$, or $v_? = v'_?$.
\end{theorem}

Note:
\begin{itemize}

  \item $\cdot \Downarrow$ is a relation between terms that is defined such that $M \Downarrow$ means
  that $M$ terminates, either with a run-time error or a value $n$ of type $\nat$.

  \item $v_?$ is shorthand for the syntactic representation of a term that is either equal to
  $\mho$, or equal to the syntactic representation of a value $n$ of type $\nat$.
\end{itemize}

% We also should be able to show that $\mho$, $\zro$, and $\suc\, N$ are not equal.

To prove graduality and validate the equational theory, we construct a model of the types
and terms and show that all of the axioms for term precision and for equality of terms
hold in this model. Modeling the dynamic type presents a challenge in the presence of a
language with functions: we want the dynamic type to represent a sum of all possible types
in the language, so we write down an recursive equation that the semantic object modeling
dynamic type should satisfy. When the language includes function types, this equation involves a
negative occurrence of the variable for which we are solving, and so the equation 
does not have inductive or coinductive solutions.
%
To model the dynamic type, we therefore use guarded recursion to define a suitable
semantic object that satisfies the unfolding isomorphism expected of the dynamic type.
The key is that we do not actually get an exact solution to the equation in the style
of traditional domain theory; rather, we get a ``guarded'' solution that holds ``up to a time step''.
%
That is, we introduce a notion of ``time'' and in the equation for the dynamic type,
we guard the negative occurrences of the variable by a special operator that
specifies that its argument is available ``later''.
%This can be seen as a logic that internalizes the notion of step-indexing.
See Section \ref{sec:technical-background} for more details on guarded type theory.

At a high level, the key parts of our proof are as follows:

% TODO revise this
\begin{itemize}
  \item Our first step toward proving graduality is to formulate a \emph{step-sensitive},
  or \emph{intensional}, gradual lambda calculus, which we call $\intlc$, in which the
  computation steps taken by a term are made explicit.
  The ``normal'' gradual lambda calculus for which we want to prove graduality will be called the
  \emph{surface}, \emph{step-insensitive}, or \emph{extensional}, gradual lambda calculus,
  denoted $\extlc$.

  \item We define a translation from the surface syntax to the intensional syntax, and
  prove a theorem relating the term precision in the surface to term precision in the
  intensional syntax.

  \item We define a semantics for the intensional syntax in guarded type theory, for both the
  terms and for the term precision ordering $\ltdyn$.

\end{itemize}

>>>>

Our goal in this work is to provide a reusable semantic framework for
gradually typed languages that can be used to prove the aforementioned
properties: graduality and type-based reasoning.

@@

The standard alternative to domain theory that scales to essentially
arbitrary self-referential definitions is \emph{step-indexing} or its
synthetic form of \emph{guarded recursion}. A series of works
\cite{new-ahmed2018, new-licata-ahmed2019, new-jamner-ahmed19}
developed step-indexed logical relations models of gradually typed
languages based on operational semantics. Unlike classical domain
theory, such step-indexed techniques are capable of modeling
higher-order store and runtime-extensible dynamic types
\cite{appelmcallester01,ahmed06,neis09,new-jamner-ahmed19}. However,
their proof developments are highly repetitive and technical, with
each development formulating a logical relation from first-principles
and proving many of the same tedious lemmas without reusable
mathematical abstractions. Our goal in the current work is to extract
these reusable mathematical principles from these explicit
step-indexed to make formalization of realistic gradual languages
tractible.

@@

An alternative approach, which we
investigate in this paper, is provided by \emph{synthetic guarded
domain theory}\cite{birkedal-mogelberg-schwinghammer-stovring2011}.
The techniques of synthetic guarded domain theory allow us to
internalize the step-indexed reasoning normally required in logical
relations proofs of graduality, ultimately allowing us to specify the
logical relation in a manner that looks nearly identical to a typical,
non-step-indexed logical relation. In fact, guarded domain theory goes
further, allowing us to define step-indexed \emph{denotational
semantics} not just step-indexed relations, just as easily as
constructing an ordinary set-theoretic semantics.

In this paper, we develop an adequate denotational semantics that
satisfies graduality and soundness of the equational theory of cast
calculi using the techniques of SGDT.  Our longer-term goal is to
mechanize these proofs in a reusable way in the Guarded Cubical Agda
proof assistant, thereby providing a framework to use to more easily
and conveniently prove that existing languages satisfy graduality and
have sound equational theories. Moreover, the aim is for designers of
new languages to utilize the framework to facilitate the design of new
provably-correct gradually-typed languages with more complex features.

@@

The main contribution of this work is a categorical and denotational
semantics for gradually typed langauges that models not just the term
language but the graduality property as well.
\begin{enumerate}
\item First, we give a simple abstract categorical model of GTT using CBPV double categories.
\item Next, we modify this semantics to develop reflexive graph- and
  double categorical models that abstract over the details of
  step-indexed models, and provide a method for constructing such models.
\item We instantiate the abstract construction to provide a concrete
  semantics in informal guarded type theory.
\item We prove that the resulting denotational model is
  \emph{adequate} for the graduality property: a closed term precision
  $M \ltdyn N : \nat$ has the expected semantics, that $M$ errors or
  $M$ and $N$ have the same extensional behavior.
\end{enumerate}

%% \subsection{Outline of Remainder of Paper}

%% \max{this is entirely out of date, update later}

%% % In Section \ref{sec:overview}, we give an overview of the gradually-typed lambda
%% % calculus and the graduality theorem.
%% %
%% In Section \ref{sec:technical-background}, we provide technical background on synthetic guarded domain theory.
%% % 
%% In Section \ref{sec:gtlc-terms}, we formally introduce the gradually-typed cast calculus
%% for which we will prove graduality. We give a semantics to the terms using
%% synthetic guarded domain theory.
%% % Important here are the notions of syntactic
%% % type precision and term precision. For reasons we describe below, we
%% % introduce two related calculi, one in which the stepping behavior of terms is
%% % implicit (an \emph{extensional} calculus, $\extlc$), and another where this behavior
%% % is made explicit (an \emph{intensional} calculus, $\intlc$).
%% %

%% In Section \ref{sec:gtlc-precision}, we define the term precision ordering for
%% the gradually-typed lambda calculus and describe our approach to assigning a
%% denotational semantics to this ordering.
%% This approach builds on the semantics constructed in the previous section,
%% but extending it to the term ordering presents new challenges.

%% % In Section \ref{sec:domain-theory}, we define several fundamental constructions
%% % internal to SGDT that will be needed when we give a denotational semantics to
%% % the gradual lambda calculus.
%% %This includes the Lift monad, predomains and error domains.
%% %
%% % In Section \ref{sec:semantics}, we define the denotational semantics for the
%% % intensional gradually-typed lambda calculus using the domain theoretic
%% % constructions defined in the previous section.
%% %
%% In Section \ref{sec:graduality}, we outline in more detail the proof of graduality for the
%% gradual lambda calculus.
%% %
%% In Section \ref{sec:discussion}, we discuss the benefits and drawbacks to our approach in comparison
%% to the traditional step-indexing approach, as well as possibilities for future work.

%% \subsection{Overview of Results}\label{sec:overview}

%% % This section used to be part of the intro.
%% % \subsection{Proving Graduality in SGDT}
%% % TODO This section should probably be moved to after the relevant background has been introduced?

%% % TODO introduce the idea of cast calculus and explicit casts?

%% In this paper, we will utilize SGDT techniques to prove graduality for a particularly
%% simple gradually-typed cast calculus, the gradually-typed lambda calculus.
%% Before stating the graduality theorem, we introduce some definitions related to gradual typing.

%% % Cast calculi
%% % TODO move this to earlier?
%% Gradually typed languages are generally elaborated to a \emph{cast calculus}, in which run-time type checking
%% that is made explicit. Elaboration inserts \emph{type casts} at the boundaries between static and dynamic code.
%% In particular, cast insertion happens at the elimination forms of the language (e.g., pattern matching, field reference, etc.).
%% The original gradually typed language that is elaborated to a cast calculus is called the \emph{surface language}.

%% % Up and down casts
%% In a cast calculus, there is a relation $\ltdyn$ on types such that $A \ltdyn B$ means that
%% $A$ is a \emph{more precise} type than $B$.
%% There a distinguished type $\dyn$, the \emph{dynamic type}, with the property that $A \ltdyn\, \dyn$ for all $A$.
%% The parts of the code that are dynamically typed will be given type $\dyn$ in the cast calculus.
%% %
%% If $A \ltdyn B$, a term $M$ of type $A$ may be \emph{up}casted to $B$, written $\up A B M$,
%% and a term $N$ of type $B$ may be \emph{down}casted to $A$, written $\dn A B N$.
%% Upcasts always succeed, while downcasts may fail at runtime, resulting in a type error.
%% Cast calculi have a term $\mho$ representing a run-time type error for any type $A$.

%% % Syntactic term precision
%% We also have a notion of \emph{syntactic term precision}.
%% If $A \ltdyn B$, and $M$ and $N$ are terms of type $A$ and $B$ respectively, we write
%% $M \ltdyn N : A \ltdyn B$ to mean that
%% $M$ is ``syntactically more precise'' than $N$, or equivalently, $N$ is 
%% ``more dynamic'' than $M$. Intuitively, this models the situation where $M$ and $N$
%% behave the same except that $M$ may error more.
%% Term precision is defined by a set of axioms that capture this intuitive notion.
%% The specific rules for term precision depend on the specifics of the language, but
%% many of the rules can be derived from the typing rules in a straightforward way.
%% % Cast calculi also have a term $\mho$ representing a run-time type error for any type $A$,
%% % and since
%% Since $\mho$ always errors, there is a term precision rule stating that $\mho \ltdyn M$ for all $M$.

%% % GTLC
%% The gradually-typed lambda calculus is the usual simply-typed lambda calculus with a dynamic
%% type $\dyn$ such that $A \ltdyn\, \dyn$ for all types $A$, as well as upcasts and downcasts
%% between any types $A$ and $B$ such that $A \ltdyn B$. The complete definition will be provided in
%% Section \ref{sec:gtlc-precision}.

%% With these definitions, we can state the graduality theorem for the gradually-typed lambda calculus.

%% % \begin{theorem}[Graduality]
%% %   If $M \ltdyn N : \nat$, then either:
%% %   \begin{enumerate}
%% %     \item $M = \mho$
%% %     \item $M = N = \zro$
%% %     \item $M = N = \suc n$ for some $n$
%% %     \item $M$ and $N$ diverge
%% %   \end{enumerate}
%% % \end{theorem}

%% \begin{theorem}[Graduality]
%%   Let $\cdot \vdash M \ltdyn N : \nat$. 
%%   If $M \Downarrow v_?$ and $N \Downarrow v'_?$, then either $v_? = \mho$, or $v_? = v'_?$.
%% \end{theorem}

%% Note:
%% \begin{itemize}

%%   \item $\cdot \Downarrow$ is a relation between terms that is defined such that $M \Downarrow$ means
%%   that $M$ terminates, either with a run-time error or a value $n$ of type $\nat$.

%%   \item $v_?$ is shorthand for the syntactic representation of a term that is either equal to
%%   $\mho$, or equal to the syntactic representation of a value $n$ of type $\nat$.
%% \end{itemize}

%% % We also should be able to show that $\mho$, $\zro$, and $\suc\, N$ are not equal.

%% To prove graduality and validate the equational theory, we construct a model of the types
%% and terms and show that all of the axioms for term precision and for equality of terms
%% hold in this model. Modeling the dynamic type presents a challenge in the presence of a
%% language with functions: we want the dynamic type to represent a sum of all possible types
%% in the language, so we write down an recursive equation that the semantic object modeling
%% dynamic type should satisfy. When the language includes function types, this equation involves a
%% negative occurrence of the variable for which we are solving, and so the equation 
%% does not have inductive or coinductive solutions.
%% %
%% To model the dynamic type, we therefore use guarded recursion to define a suitable
%% semantic object that satisfies the unfolding isomorphism expected of the dynamic type.
%% The key is that we do not actually get an exact solution to the equation in the style
%% of traditional domain theory; rather, we get a ``guarded'' solution that holds ``up to a time step''.
%% %
%% That is, we introduce a notion of ``time'' and in the equation for the dynamic type,
%% we guard the negative occurrences of the variable by a special operator that
%% specifies that its argument is available ``later''.
%% %This can be seen as a logic that internalizes the notion of step-indexing.
%% See Section \ref{sec:technical-background} for more details on guarded type theory.

%% At a high level, the key parts of our proof are as follows:

%% % TODO revise this
%% \begin{itemize}
%%   \item Our first step toward proving graduality is to formulate a \emph{step-sensitive},
%%   or \emph{intensional}, gradual lambda calculus, which we call $\intlc$, in which the
%%   computation steps taken by a term are made explicit.
%%   The ``normal'' gradual lambda calculus for which we want to prove graduality will be called the
%%   \emph{surface}, \emph{step-insensitive}, or \emph{extensional}, gradual lambda calculus,
%%   denoted $\extlc$.

%%   \item We define a translation from the surface syntax to the intensional syntax, and
%%   prove a theorem relating the term precision in the surface to term precision in the
%%   intensional syntax.

%%   \item We define a semantics for the intensional syntax in guarded type theory, for both the
%%   terms and for the term precision ordering $\ltdyn$.

%% \end{itemize}

### 2024-01-27 paper-new/technical-background.tex
Revise sections 1 and 2

\section{Technical Background}\label{sec:technical-background}

@@

\subsection{Difficulties in Prior Semantics}
  % Difficulties in prior semantics

  In this work, we compare our approach to proving graduality to the approach
  introduced by New and Ahmed \cite{new-ahmed2018} which constructs a step-indexed
  logical relations model and shows that this model is sound with respect to their
  notion of contextual error approximation.

  Because the dynamic type is modeled as a non-well-founded
  recursive type, their logical relation needs to be paramterized by natural numbers
  to restore well-foundedness. This technique is known as a \emph{step-indexed logical relation}.
  Intuitively, step-indexing makes the steps taken by terms observable.
  %
  Reasoning about step-indexed logical relations
  can be tedious and error-prone, and there are some very subtle aspects that must
  be taken into account in the proofs.
  %
  In particular, the prior approach of New and Ahmed requires two separate logical
  relations for terms, one in which the steps of the left-hand term are counted,
  and another in which the steps of the right-hand term are counted.
  Then two terms $M$ and $N$ are related in the ``combined'' logical relation if they are
  related in both of the one-sided logical relations. Having two separate logical relations
  complicates the statement of the lemmas used to prove graduality, because any statement that
  involves a term stepping needs to take into account whether we are counting steps on the left
  or the right. Some of the differences can be abstracted over, but difficulties arise for properties %/results
  as fundamental and seemingly straightforward as transitivity.

  Specifically, for transitivity, we would like to say that if $M$ is related to $N$ at
  index $i$ and $N$ is related to $P$ at index $i$, then $M$ is related to $P$ at $i$.
  But this does not actually hold: we require that one of the two pairs of terms
  be related ``at infinity'', i.e., that they are related at $i$ for all $i \in \mathbb{N}$.
  Which pair is required to satisfy this depends on which logical relation we are considering,
  (i.e., is it counting steps on the left or on the right),
  and so any argument that uses transitivity needs to consider two cases, one
  where $M$ and $N$ must be shown to be related for all $i$, and another where $N$ and $P$ must
  be related for all $i$. % This may not even be possible to show in some scenarios!

  \begin{comment}
  As another example, the axioms that specify the behavior of casts do not hold in the
  step-indexed setting. Consider, for example, the ``lower bound'' rule for downcasting:

  \begin{mathpar}
    \inferrule*
    {\gamlt \vdash M \ltdyn N : dyn}
    {\gamlt \vdash \dn{\dyn \ra \dyn}{\dyn} M \ltdyn N}
  \end{mathpar}

  In the language of the step-indexed logical relation used in prior work, this would
  take the form

  \begin{mathpar}
    \inferrule*
    {(M, N) \in \mathcal{E}^{\sim}_{i}\sem{\dyn}}
    {(\dn{\dyn \ra \dyn}{\dyn} M, N) \in \mathcal{E}^{\sim}_{i}\sem{\injarr{}}}
  \end{mathpar}

  where $\sim$ stands for $\mathrel{\preceq}$ or $\mathrel{\succeq}$, i.e.,
  counting steps on the left or right respectively.
  %
  To show this, we would use the fact that the left-hand side steps and
  apply an anti-reduction lemma, showing that the term to which the LHS steps
  is related to the RHS where our fuel is now

  The left-hand side steps to a case inspection on $M$,
  where we unfold the recursive $\dyn$ type into a sum type and see whether the result
  is a function type.

  One way around these difficulties is to demand that the rules only hold
  ``extensionally'', i.e., we quantify universally over the step-index and
  reason about the ``global'' behavior of terms for all possible step indices.
  This is the approach taken in prior work.
\end{comment}

  % These complications introduced by step-indexing lead one to wonder whether there is a
  % way of proving graduality without relying on tedious arguments involving natural numbers.
  % An alternative approach, which we investigate in this paper, is provided by
  % \emph{synthetic guarded domain theory}, as discussed below.
  % Synthetic guarded domain theory allows the resulting logical relation to look almost
  % identical to a typical, non-step-indexed logical relation.

\subsection{Synthetic Guarded Domain Theory}\label{sec:sgdt}
One way to avoid the tedious reasoning associated with step-indexing is to work
axiomatically inside of a logical system that can reason about non-well-founded recursive
constructions while abstracting away the specific details of step-indexing required
if we were working analytically.
The system that proves useful for this purpose is called \emph{synthetic guarded
domain theory}, or SGDT for short. We provide a brief overview here, but more
details can be found in \cite{birkedal-mogelberg-schwinghammer-stovring2011}.

@@

% Clocked Cubical Type Theory
\subsubsection{Ticked Cubical Type Theory}
% TODO motivation for Clocked Cubical Type Theory, e.g., delayed substitutions?

>>>>

\section{Background on Guarded Domain Theory}\label{sec:technical-background}

@@

%% \subsection{Difficulties in Prior Semantics}
%%   % Difficulties in prior semantics

%%   In this work, we compare our approach to proving graduality to the approach
%%   introduced by New and Ahmed \cite{new-ahmed2018} which constructs a step-indexed
%%   logical relations model and shows that this model is sound with respect to their
%%   notion of contextual error approximation.

%%   Because the dynamic type is modeled as a non-well-founded
%%   recursive type, their logical relation needs to be paramterized by natural numbers
%%   to restore well-foundedness. This technique is known as a \emph{step-indexed logical relation}.
%%   Intuitively, step-indexing makes the steps taken by terms observable.
%%   %
%%   Reasoning about step-indexed logical relations
%%   can be tedious and error-prone, and there are some very subtle aspects that must
%%   be taken into account in the proofs.
%%   %
%%   In particular, the prior approach of New and Ahmed requires two separate logical
%%   relations for terms, one in which the steps of the left-hand term are counted,
%%   and another in which the steps of the right-hand term are counted.
%%   Then two terms $M$ and $N$ are related in the ``combined'' logical relation if they are
%%   related in both of the one-sided logical relations. Having two separate logical relations
%%   complicates the statement of the lemmas used to prove graduality, because any statement that
%%   involves a term stepping needs to take into account whether we are counting steps on the left
%%   or the right. Some of the differences can be abstracted over, but difficulties arise for properties %/results
%%   as fundamental and seemingly straightforward as transitivity.

%%   Specifically, for transitivity, we would like to say that if $M$ is related to $N$ at
%%   index $i$ and $N$ is related to $P$ at index $i$, then $M$ is related to $P$ at $i$.
%%   But this does not actually hold: we require that one of the two pairs of terms
%%   be related ``at infinity'', i.e., that they are related at $i$ for all $i \in \mathbb{N}$.
%%   Which pair is required to satisfy this depends on which logical relation we are considering,
%%   (i.e., is it counting steps on the left or on the right),
%%   and so any argument that uses transitivity needs to consider two cases, one
%%   where $M$ and $N$ must be shown to be related for all $i$, and another where $N$ and $P$ must
%%   be related for all $i$. % This may not even be possible to show in some scenarios!

%%   \begin{comment}
%%   As another example, the axioms that specify the behavior of casts do not hold in the
%%   step-indexed setting. Consider, for example, the ``lower bound'' rule for downcasting:

%%   \begin{mathpar}
%%     \inferrule*
%%     {\gamlt \vdash M \ltdyn N : dyn}
%%     {\gamlt \vdash \dn{\dyn \ra \dyn}{\dyn} M \ltdyn N}
%%   \end{mathpar}

%%   In the language of the step-indexed logical relation used in prior work, this would
%%   take the form

%%   \begin{mathpar}
%%     \inferrule*
%%     {(M, N) \in \mathcal{E}^{\sim}_{i}\sem{\dyn}}
%%     {(\dn{\dyn \ra \dyn}{\dyn} M, N) \in \mathcal{E}^{\sim}_{i}\sem{\injarr{}}}
%%   \end{mathpar}

%%   where $\sim$ stands for $\mathrel{\preceq}$ or $\mathrel{\succeq}$, i.e.,
%%   counting steps on the left or right respectively.
%%   %
%%   To show this, we would use the fact that the left-hand side steps and
%%   apply an anti-reduction lemma, showing that the term to which the LHS steps
%%   is related to the RHS where our fuel is now

%%   The left-hand side steps to a case inspection on $M$,
%%   where we unfold the recursive $\dyn$ type into a sum type and see whether the result
%%   is a function type.

%%   One way around these difficulties is to demand that the rules only hold
%%   ``extensionally'', i.e., we quantify universally over the step-index and
%%   reason about the ``global'' behavior of terms for all possible step indices.
%%   This is the approach taken in prior work.
%% \end{comment}

%%   % These complications introduced by step-indexing lead one to wonder whether there is a
%%   % way of proving graduality without relying on tedious arguments involving natural numbers.
%%   % An alternative approach, which we investigate in this paper, is provided by
%%   % \emph{synthetic guarded domain theory}, as discussed below.
%%   % Synthetic guarded domain theory allows the resulting logical relation to look almost
%%   % identical to a typical, non-step-indexed logical relation.

%% \subsection{Synthetic Guarded Domain Theory}\label{sec:sgdt}
One way to avoid the tedious reasoning associated with step-indexing is to work
axiomatically inside of a logical system that can reason about non-well-founded recursive
constructions while abstracting away the specific details of step-indexing required
if we were working analytically.
The system that proves useful for this purpose is called \emph{synthetic guarded
domain theory}, or SGDT for short. We provide a brief overview here, but more
details can be found in \cite{birkedal-mogelberg-schwinghammer-stovring2011}.

@@

% Clocked Cubical Type Theory
\subsection{Ticked Cubical Type Theory}
% TODO motivation for Clocked Cubical Type Theory, e.g., delayed substitutions?
